Optimal updating magnitude in adaptive flat-distribution-sampling simulations

### 2017-06-24 doc/invt.tex
invt.tex: improve massq, V6.3

\documentclass[reprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{In
going to the continuous time setup,
we find it more convenient to shift the origin of time by $-1$,
e.g. the sum $\sum_{i=1}^T$ is mapped to the integral $\int_0^T dt$.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { \rho_i }
  \equiv
  \alpha(t) \, f_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined
$f_i(t) \equiv h_i(t) /\rho_i$.
%
Clearly,
%
\begin{equation}
  \bav{f}(t) = \sum_{i=1}^n h_i(t) = 1
  ,
  \label{eq:fav1}
\end{equation}
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, f_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$ steps.
%
If the starting bias potential is zero,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \vartheta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \vartheta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{}
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
This serves as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define
a normalized mass distribution as
%
\begin{equation}
  m(\bar q)
  =
  \frac{
    M(\bar q)
  }
  {
    \int_0^{ q(T) } M(q') \, d q'
  }
  =
  \frac{
    M(\bar q)
  }
  {
    C_M
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(\bar q) \, d\bar q = 1$.
%
Then from Eqs. \eqref{eq:Lagrangian_const} and \eqref{eq:q_opt},
we find that for $\bar q(t) \equiv q(T) - q(t)$,
%
\begin{align}
  m\left( \bar q(t) \right)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
  \label{eq:mQ_invTa}
  \\
  \int_{\bar q(t)}^{ q(T) }
    m(q') \, d q'
  &=
  \frac t T
  ,
  \label{eq:intmQ_tT}
\end{align}
%
Thus, each point on the schedule,
$\bigl(t, \alpha(t)\bigr)$,
can be mapped to a point,
$\Bigl(\bar q(t), m\bigl(\bar q(t)\bigr)\Bigr)$,
on the mass distribution,
such that the ordinate of the latter curve
is $1/(T\alpha)$,
and the area under the curve in the domain, $[\bar q(t), q(T)]$,
is equal to $t/T$,
as shown in Fig. \ref{fig:massq}.

\begin{figure}[h]
\begin{center}
  \begin{tikzpicture}[very thick, scale=2.5, every node/.style={scale=1.5}]
    % distribution
    \draw[domain=0:2.5, variable=\x, dashed, gray, ultra thick]
      plot({\x}, {sqrt(2*exp(-2*\x)+exp(-\x))});

    % shade
    \fill [gray!30!white, domain=0.7:2.2, variable=\x]
      (0.7, 0) --
      plot ({\x}, {sqrt(2*exp(-2*\x)+exp(-\x))})
      -- (2.2, 0) -- cycle;

    % thick curve
    \draw[domain=0:2.2, variable=\x, ultra thick]
      plot({\x}, {sqrt(2*exp(-2*\x)+exp(-\x))});

    % label for the shaded area
    \node[] at (1.1, 0.40) {$\dfrac{t}{T}$};

    \draw[<->] (0.6, 0)  --
               node[left, inner sep=1mm]
               {$\dfrac{1}{T \, \alpha}$}
               (0.6, 0.99488);
    % label for the ordinate
    %\node[fill=white] at (0.4, 0.7) {$\dfrac{1}{Ta}$};

    \draw[] (0.5, 0.99488) -- (0.7, 0.99488);
    \draw[->] (0, 0) -- (2.8, 0);
    \node[] at (2.8, -0.2) {$\bar q$};
    \draw[->] (0, 0) -- node[above, rotate=90] {$m(\bar q)$} (0, 1.9);

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (0.7, 0.1) -- (0.7, 0.0) node[below] {$q(T) - q(t)$};
    \draw[] (2.2, 0.1) -- (2.2, 0.0) node[below] {$q(T)$};
  \end{tikzpicture}
  %\makebox[\linewidth][c]{
  %  \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  %}
  \caption{
    \label{fig:massq}
    Geometric construction of the optimal schedule,
    $\alpha(t)$,
    from the normalized mass distribution,
    $m(\bar q)$.
    %
    \note{
      The figure was produced by \texttt{tikz}.
      % Gnuplot version by doc/fig/massq.gp
    }%
  }
\end{center}
\end{figure}

We may consider the schedule, $\alpha(t)$,
as a variable transformation of
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited in $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$.
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution is a single exponential decay,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,\footnote{If
  the magntidue of the eigenvalues, $\lambda_k$'s,
  span a wide range, then for a large $\bar q$,
  the sum in Eq. \eqref{eq:mass_func} will be dominated by
  the largest term, where $\lambda_k \approx 1/\bar q$,
  and $M(\bar q) \approx \sqrt{\Gamma_k}/\bar q$.
  }
  %
then we would have an exponentially decaying schedule.

\begin{table}[h]
  \caption{\label{tab:m_and_a}
    Examples of the variable transformation between the
    reduced schedule, $\alpha^*(t^*)$,
    and the reduced mass distribution, $m^*(\bar q^*)$,
    as defined in Eqs. \eqref{eq:mQ_invTa}
    and \eqref{eq:intmQ_tT}.
  }
  \setlength{\tabcolsep}{3pt} % spacing between columns
  \renewcommand\arraystretch{2.0} % spacing between rows
  \begin{tabular} { c c c }
    \hline
    $m^*(\bar q^*)$ &
    $\alpha^*(t^*)$ &
    $\bar q^*(t^*)$
    \\
    \hline
    $\dfrac{1}{\gamma \, ({\bar q}^* + s_\gamma) }$ &
    $(1 + s_\gamma) \, \gamma \, e^{-\gamma \, t^*}$ &
    $(1 + s_\gamma) \, e^{-\gamma \, t^*} - s_\gamma$
    \\
    $(1+ s_\gamma) \, \gamma \, e^{-\gamma \, {\bar q}^*}$ &
    $\dfrac{1}{ \gamma \, (t^* + s_\gamma) }$ &
    $\dfrac{1}{\gamma} \ln\left(\dfrac{1 + s_\gamma}{t^* + s_\gamma} \right)$
    \\
    $\dfrac{c \, a}{\left( {\bar q}^* + q^*_0 \right)^{a+1}}$ &
    $\dfrac{c^{1/a} \, (1/a)}{\left( t^* + t^*_0 \right)^{(1/a)+1}}$ &
    %$\left( \dfrac{c}{t^* + t^*_0} \right)^{1/a} - q^*_0$
    $q^*_0 \left[\left( \dfrac{1+t^*_0}{t^* + t^*_0} \right)^{\frac 1 a} - 1\right]$
    %$\,\begin{aligned}
    %  &a, q^*_0 > 0, \\
    %  &\frac{1}{t^*_0} \equiv \left(1+\frac{1}{q^*_0}\right)^a - 1, \\
    %  &c \equiv {q^*_0}^a(t^*_0 + 1)
    %\end{aligned}$
    \\
    \hline
    \multicolumn{3}{p{8cm}}{
    Here,
    $\gamma > 0$,
    and we have defined
    $s_\gamma \equiv 1/(e^\gamma - 1)$,
    ${t^*_0}^{-1} \equiv \left(1+{q^*_0}^{-1}\right)^a - 1$,
    and
    $c \equiv {q^*_0}^a(1 + t^*_0)$.
    } \\
    \hline
  \end{tabular}
\end{table}

@@

%For simulations with the same $a_0$,
%the shape of the mass distribution, $m(\bar q)$,
%also determines how much the optimal schedule
%would depend on the simulation length, $T$.
%%
%Since both $M(\bar q)$ and $m(\bar q)$ decrease monotonically with $\bar q$,
%Eq. \eqref{eq:opt_qT} shows that
%the optimal $q(T)$ increases with
%the simulation length, $T$,
%more so if $M(\bar q)$ has a long tail.
%%
%Thus, if $M(\bar q)$ decays slowly because of
%many near zero eigenvalues, $\lambda_k$'s,
%as in the metadynamics case,
%both the optimal schedule and
%normalized asymptotic error, $\Err_A(T) \cdot T$,
%would be sensitive to the simulation length, $T$.
%%
%In contrast, if all eigenvalues are $1.0$
%as in the single-bin or WL updating scheme discussed previously,
%the optimal schedule is
%given by the universal inverse-time formula,
%Eq. \eqref{eq:alpha_invt1},
%and the normalized asymptotic error
%is roughly a constant
%[cf. Eq. \eqref{eq:error_singlebin}].
%
%If $M(\bar q)$ decays rapidly,
%the distribution $m(\bar q)$ does not critically
%depend on the cutoff $q(T)$,
%once $q(T)$ is sufficiently large.
%%
%So the optimal schedules of these simulations
%approximately satisfy the scaling relations
%described in Sec. \ref{sec:mass_distr}.
%%
%If, however, $M(\bar q)$ has a long tail
%because of some near-zero $\lambda_k$'s,
%increasing the simulation length $T$
%generally alters the shape of the optimal schedule.

@@

Based on the above results,
we now give a procedure of computing
the optimal schedule and the error
for a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_k$'s and $\Gamma_k$'s
from the given system and updating scheme (input),
and then apply them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process.
%
We have

@@

%\subsubsection{\label{sec:bandkernel}
%Updating kernel and eigenvalues}

>>>>

\documentclass[preprint, superscriptaddress, floatfix]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage[table,usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{hyperref}
\usepackage{setspace}

@@

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{In
going to the continuous-time setup,
we find it more convenient to shift the origin of time by $-1$,
e.g. the sum $\sum_{i=1}^T$ is mapped to the integral $\int_0^T dt$.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { \rho_i }
  \equiv
  \alpha(t) \, f_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined
$f_i(t) \equiv h_i(t) /\rho_i$.
%
Clearly,
%
\begin{equation}
  \bav{f}(t) = \sum_{i=1}^n h_i(t) = 1
  ,
  \label{eq:fav1}
\end{equation}
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, f_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the preliminary adaptive FDS sampling has been
performed under a constant updating magnitude for $T_0$ steps.
%
If the initial bias potential
before the priliminary run is zero,
one can show that
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  + \epsilon_k
  ,
  \label{eq:xt2_eql1}
\end{equation}
%
where
$\epsilon_k \equiv \tilde u_k^2 \, e^{-2\, \lambda_k \, a_0 \, T_0}$
and if no eigenvalue is zero,
then we can ignore the $\epsilon_k$ term
in the case of $T_0 \to \infty$,
%
\begin{equation}
  \left\langle
    {\tilde v}_k^2(0)
  \right\rangle
  = \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k
  .
  \label{eq:xt2_eql}
\end{equation}
%
\note{
Consider a long period, $T_0$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle
    {\tilde v}_k^2(T_0)
  \right\rangle
  =
  \int_0^{T_0}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T_0) }
    \, dt
  \stackrel{ T_0 \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
  \notag
  %\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot \vartheta}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, \vartheta_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(\bar q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}
    \Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, \bar q}
  }
  .
  %\notag
  \label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,\cite{}
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C_M}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C_M$.
%
This serves as a generalization of Eq. \eqref{eq:dthetadt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(\bar q)
    \;
    d \bar q
  =
  C_M \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C_M$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C_M =
  \int_{ 0 }^{ q(T) }
    M( \bar q )
    \;
    d \bar q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C_M^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(\bar q) \, d \bar q
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}

@@

We can characterize the optimal schedule
geometrically as follows.
%
First, we define
a normalized mass distribution as
%
\begin{equation}
  m(\bar q)
  =
  \frac{
    M(\bar q)
  }
  {
    \int_0^{ q(T) } M(q') \, d q'
  }
  =
  \frac{
    M(\bar q)
  }
  {
    C_M
  }
  ,
\notag
%\label{eq:mass_distr}
\end{equation}
%
such that
$\int_0^{q(T)} m(\bar q) \, d\bar q = 1$.
%
Then from Eqs. \eqref{eq:Lagrangian_const} and \eqref{eq:q_opt},
we find that for $\bar q(t) \equiv q(T) - q(t)$,
%
\begin{align}
  m\left( \bar q(t) \right)
  &=
  \frac{ 1 }
       { T \, \alpha(t) }
  ,
  \label{eq:mQ_invTa}
  \\
  \int_{\bar q(t)}^{ q(T) }
    m(q') \, d q'
  &=
  \frac t T
  ,
  \label{eq:intmQ_tT}
\end{align}
%
Thus, each point on the schedule,
$\bigl(t, \alpha(t)\bigr)$,
can be mapped to a point,
$\Bigl(\bar q(t), m\bigl(\bar q(t)\bigr)\Bigr)$,
on the mass distribution,
such that the ordinate of the latter curve
is $1/(T\alpha)$,
and the area under the curve in the domain, $[\bar q(t), q(T)]$,
is equal to $t/T$,
as shown in Fig. \ref{fig:massq}(a).

\begin{figure}[h]
\begin{center}
  \begin{tikzpicture}[thick, scale=1.25, every node/.style={scale=0.8}]
    \node [] at (-0.3, 2.0) {(a)};
    % distribution
    \draw[domain=0:2.5, variable=\x, dashed, gray, very thick]
      plot({\x}, {2*exp(-\x)});

    % shade
    \fill [gray!30!white, domain=0.7:2.2, variable=\x]
      (0.7, 0) --
      plot ({\x}, {2*exp(-\x)})
      -- (2.2, 0) -- cycle;

    % thick curve
    \draw[domain=0:2.2, variable=\x, very thick]
      plot({\x}, {2*exp(-\x)});

    % label for the shaded area
    \node[] at (1.1, 0.36) {$\dfrac{t}{T}$};

    \draw[<->] (0.6, 0)  --
               node[left, inner sep=1mm]
               {$\dfrac{1}{T \, \alpha}$}
               (0.6, 0.99317);
    % label for the ordinate
    %\node[fill=white] at (0.4, 0.7) {$\dfrac{1}{Ta}$};

    \draw[] (0.5, 0.99317) -- (0.7, 0.99317);
    \draw[->] (0, 0) -- (2.8, 0);
    \node[] at (2.8, -0.2) {$\bar q$};
    \draw[->] (0, 0) -- node[above, rotate=90] {$m(\bar q)$} (0, 2.2);

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (0.7, 0.1) -- (0.7, 0.0) node[below] {$q(T) - q(t)$};
    \draw[] (2.2, 0.1) -- (2.2, 0.0) node[below] {$q(T)$};
  \end{tikzpicture}
%
  \begin{tikzpicture}[thick, scale=1.25, every node/.style={scale=0.8}]
    \node [] at (-0.3, 2.0) {(b)};
    % distribution
    \draw[domain=0:2.3, variable=\x, dashed, gray, very thick]
      plot({\x}, {1/(\x+0.5)});

    % shade
    \fill [gray!30!white, domain=0.7:2.0, variable=\x]
      (0.7, 0) --
      plot ({\x}, {1/(\x+0.5)})
      -- (2.0, 0) -- cycle;

    % thick curve
    \draw[domain=0:2.0, variable=\x, very thick]
      plot({\x}, {1/(\x+0.5)});

    % label for the shaded area
    \node[] at (1.1, 0.40) {$\bar q(t)$};

    \draw[<->] (0.6, 0)  --
               node[left, inner sep=1mm]
               {$\dfrac{1}{T \, m}$}
               (0.6, 0.83333);
    % label for the ordinate
    %\node[fill=white] at (0.4, 0.7) {$\dfrac{1}{Ta}$};

    \draw[] (0.5, 0.83333) -- (0.7, 0.83333);
    \draw[->] (0, 0) -- (2.6, 0);
    \node[] at (2.6, -0.2) {$t$};
    \draw[->] (0, 0) -- node[above, rotate=90] {$\alpha(t)$} (0, 2.2);

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (0.7, 0.1) -- (0.7, 0.0) node[below] {$t$};
    \draw[] (2.0, 0.1) -- (2.0, 0.0) node[below] {$T$};
  \end{tikzpicture}
  \caption{
    \label{fig:massq}
    (a) Geometric construction of the optimal schedule,
    $\alpha(t)$,
    from the normalized mass distribution,
    $m(\bar q)$.
    %
    (b) Inverse construction of $m(\bar q)$
    from the optimal schedule $\alpha(t)$.
    %
    \note{
      The figure was produced by \texttt{tikz}.
    }%
  }
\end{center}
\end{figure}

We may consider the schedule, $\alpha(t)$,
as a variable transformation of
the mass distribution, $m(\bar q)$.
%much like the Legendre transform.
%
If we define
$\bar q^* \equiv \bar q/q(T)$
and $t^* \equiv t/T$
such that both are limited in $[0, 1]$,
as well as
\begin{align*}
  m^*       &\equiv q(T) \, m(\bar q) = -\frac{ dt^* } { d\bar q^* },
  \\
  \alpha^*  &\equiv \frac{T}{q(T)} \, \alpha(t) = -\frac{ d\bar q^* } { dt^* },
\end{align*}
then it is clear that
both $m^*(\bar q^*)$ and $\alpha^*(t^*)$
can describe the differential relation
between $\bar q^*$ and $t^*$,
but with the natural variables
being $\bar q^*$ and $t^*$, respectively.
%
Further, the transformation is an involution:
if we apply the same transform on $\alpha^*(t^*)$
we would recover $m^*(\bar q^*)$,
as shown in Fig. \ref{fig:massq}.
%
A few examples are shown in Table \ref{tab:m_and_a}.
%
For example, in the single-bin scheme,
the mass distribution is a single exponential decay,
and we would have the inverse-time schedule.
%
On the other hand,
if the mass distribution is inversely proportional to $\bar q$,
which is approximately true for the Gaussian updating scheme
at large $\bar q$,\footnote{If
  the magnitude of the eigenvalues, $\lambda_k$'s,
  span a wide range, then for a large $\bar q$,
  the sum in Eq. \eqref{eq:mass_func} will be dominated by
  the largest term, where $\lambda_k \approx 1/\bar q$,
  and $M(\bar q) \approx \sqrt{\Gamma_k}/\bar q$.
  }
  %
then we would have an exponentially decaying schedule.

\begin{table}[h]
  \caption{\label{tab:m_and_a}
    Examples of the variable transformation between the
    reduced schedule, $\alpha^*(t^*)$,
    and the reduced mass distribution, $m^*(\bar q^*)$,
    as defined in Eqs. \eqref{eq:mQ_invTa}
    and \eqref{eq:intmQ_tT}.
  }
  \setlength{\tabcolsep}{2pt} % spacing between columns
  \renewcommand\arraystretch{2.0} % spacing between rows
  \begin{tabular} { c c c }
    \hline
    $m^*(\bar q^*)$ &
    $\alpha^*(t^*)$ &
    $\bar q^*(t^*)$
    \\
    \hline
    $\dfrac{1}{\gamma \, ({\bar q}^* + s_\gamma) }$ &
    $(1 + s_\gamma) \, \gamma \, e^{-\gamma \, t^*}$ &
    $(1 + s_\gamma) \, e^{-\gamma \, t^*} - s_\gamma$
    \\
    $(1+ s_\gamma) \, \gamma \, e^{-\gamma \, {\bar q}^*}$ &
    $\dfrac{1}{ \gamma \, (t^* + s_\gamma) }$ &
    $\dfrac{1}{\gamma} \ln\left(\dfrac{1 + s_\gamma}{t^* + s_\gamma} \right)$
    \\
    $\dfrac{c \, a}{\left( {\bar q}^* + q^*_0 \right)^{a+1}}$ &
    $\dfrac{c^{1/a} \, (1/a)}{\left( t^* + t^*_0 \right)^{(1/a)+1}}$ &
    %$\left( \dfrac{c}{t^* + t^*_0} \right)^{1/a} - q^*_0$
    $q^*_0 \left[\left( \dfrac{1+t^*_0}{t^* + t^*_0} \right)^{1/a} - 1\right]$
    %$\,\begin{aligned}
    %  &a, q^*_0 > 0, \\
    %  &\frac{1}{t^*_0} \equiv \left(1+\frac{1}{q^*_0}\right)^a - 1, \\
    %  &c \equiv {q^*_0}^a(t^*_0 + 1)
    %\end{aligned}$
    \\
    \hline
    \multicolumn{3}{p{8cm}}{
    Here,
    $\gamma > 0$,
    and we have defined
    $s_\gamma \equiv 1/(e^\gamma - 1)$,
    ${t^*_0}^{-1} \equiv \left(1+{q^*_0}^{-1}\right)^a - 1$,
    and
    $c \equiv {q^*_0}^a(1 + t^*_0)$.
    } \\
    \hline
  \end{tabular}
\end{table}

@@

@@

Based on the above results,
we now give a procedure of computing
the optimal schedule and the error
for a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_k$'s and $\Gamma_k$'s
from the given system and updating scheme (input),
and then apply them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process.
%
We have

@@
